\chapter{Implementation and Methodology}

\section{Implementation Goals}

The implementation part of this thesis was designed with two main goals. The
first was to create a working experimental environment for the Merkle--Hellman
knapsack cryptosystem and selected related variants. The second was to provide
a practical basis for comparing their behaviour, especially with respect to key
generation, encryption, decryption, and later cryptanalytic experiments.
For this reason, the implementation was not intended as a production-ready
cryptographic library. Instead, it was developed as a small research-oriented
program in which the main algorithms can be tested, modified, and benchmarked
in a controlled way. This made it possible to focus on the structure of the
schemes themselves and on the consequences of different design choices, such as
the use of permutation or repeated transformations.
Another important goal was reproducibility. The program supports deterministic
initialization through explicit seeds, which makes it possible to regenerate
the same keys and benchmark scenarios across multiple runs. This is useful both
for debugging and for experimental evaluation in later chapters.

\section{Choice of Programming Language and Libraries}

The implementation was written in the C programming language. This choice was
made mainly for practical reasons. C provides low-level control over memory and
data representation, introduces little abstraction overhead, and allows the
individual steps of key generation, encryption, and decryption to be expressed
in a direct way. Since the implemented schemes are mathematically simple but
operate on potentially large integers, this level of control was useful for
both debugging and benchmarking.
At the same time, native integer types are not sufficient for realistic
knapsack computations, because the weights, modulus, multiplier, and ciphertext
can quickly exceed fixed machine word sizes. For this reason, the
implementation uses the GNU Multiple Precision Arithmetic Library (GMP). GMP is
used for the core arithmetic operations, especially modular multiplication,
modular inversion, and ciphertext computation. This made it possible to keep
the implementation in C while still supporting arithmetic beyond standard
64-bit limits.
The build process is controlled by a simple Makefile. It supports separate
debug and release builds, as well as an address-sanitized configuration for
runtime checking. This was useful during development, because the code makes
extensive use of dynamic allocation and scheme-specific internal structures.
The overall goal was to keep the toolchain simple, portable, and suitable for
experimental work rather than for deployment.

\section{Program Architecture}

The implementation was organized as a small modular C program. The main goal
was to keep the cryptographic code separate from user interaction, input
parsing, and benchmarking. This made the program easier to test and also made
it possible to add new knapsack variants without changing the rest of the
application structure.

\subsection{Top-Level Structure}
The executable entry point in `main.c` is intentionally minimal and delegates
control to the application layer in `app.c`. This top-level layer is
responsible for the overall execution flow, including mode selection, demo
execution, plaintext handling, and benchmark runs. Command-line parsing and
validation are centralized in `cli.c`, which keeps input handling separate from
the cryptographic part of the program.
The program currently supports two main execution modes. The first is a demo
mode intended for direct experiments with encryption and decryption. The second
is a benchmark mode, which repeatedly runs the selected scheme and records
timing data for later evaluation. This separation made it possible to keep the
interactive and experimental parts of the implementation independent.

\subsection{Scheme Abstraction}
At the core of the implementation is a small scheme abstraction defined in
`scheme.h`. Each cryptosystem variant is represented by a `SchemeOps`
structure containing function pointers for key generation, encryption,
decryption, and cleanup, together with a short identifier and a display name.
This provides a common interface for all implemented schemes while still
allowing each variant to keep its own internal data structures.
The currently implemented variants are the classical Merkle--Hellman scheme and
a permuted variant. Their implementations are placed in separate source files,
which keeps the code easier to read and compare. Scheme selection is handled in
`scheme.c`, where the requested identifier is resolved to the corresponding
implementation.

\subsection{Data Representation and Support Modules}
Messages are represented internally as bit vectors through the `BitView` and
`BitBuf` data structures. This matches the block-based nature of the knapsack
schemes and makes it straightforward to work with fixed-length binary message
blocks. For plaintext demonstrations, the helper logic converts text into bit
blocks before encryption and reconstructs the original text after decryption.
Several supporting modules are used around the core scheme implementations.
Seed handling and entropy acquisition are separated into `seed.c`, which allows
both deterministic runs and randomly initialized ones. Benchmarking is handled
in `bench.c`, where key generation, encryption, and decryption are measured
separately. Correctness is verified by checking that the decrypted output
matches the original message. In this way, the overall architecture remains
small, but still clearly separates cryptographic operations, application flow,
and experimental evaluation.

\section{Testing and Verification}
% Correctness, performance

