\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}

Cryptography is a fundamental part of modern information security. It enables
confidential communication by transforming data into forms that are
unaccessible to unauthorized parties. In today’s digital world, cryptography is
is being used in almost every secure communication system, from online
banking and data storage to digital signatures and blockchain technologies. Its
roots go back thousands of years. From simple substitution ciphers like Ceasar
cipher used in ancient civilizations, through more complex systems like the
Vigenère cipher, to the formal mathematical foundations developed in the
twentieth century.  

%The main goal of
% cryptography has always been to ensure confidentiality, integrity, and
% authenticity of information.

Throughout it's history cryptography reliad mostly on symmetric cryptosystems,
where both the sender and reciever use the same private key for both encryption
and decryption. While these systems, such as DES and AES, are efficient and still
widely used today, they require both parties to securely share the secret key in
advance. Which comes with it's own problems

A major milestone in the history of cryptography came with the invention of
public-key cryptography in the 1970s. Unlike classical symmetric systems, where
both parties share the same secret key, public-key systems use a pair of
mathematically linked keys. Public key for encryption and a private key for
decryption. This enables secure communication between parties without
exchanging keys before communication.  Among the most well-known public-key systems
are the RSA cryptosystem, based on the difficulty of integer factorization, and
the Diffie–Hellman key exchange, which relies on the discrete logarithm
problem.

Alongside the first systems such as RSA, one of the earliest public-key systems
was the Knapsack cryptosystem proposed by Merkle and Hellman in 1978, which
relied on the computational difficulty of the Knapsack problem, a classic
example of an NP-complete combinatorial problem. The Knapsack problem asks
whether a given sum can be formed by selecting specific elements from a set of
numbers, and while it is easy to verify a solution, finding it is
computationally hard. This asymmetry between easy verification and hard
computation makes the problem appealing for cryptographic use. 

Although the original Merkle–Hellman system was later shown to be insecure
against certain attacks, it remains an important milestone in cryptographic
research. Its concept has inspired numerous modifications and improved versions
that attempt to overcome the weaknesses of the original design. Studying these
variations provides valuable insight into both the design principles of secure
systems and the evolution of cryptographic thinking.

The aim of this thesis is to study, analyze, compare, and implement various
versions of the Knapsack cryptosystem. The work begins with an overview of
basic cryptographic concepts and the theoretical background of the Knapsack
problem. It then describes the classical Merkle–Hellman Knapsack cryptosystem
in detail, including its key generation, encryption, and decryption procedures.
The following sections analyze its security, explain known cryptanalytic
attacks, and present selected improvements and modern adaptations. Finally, an
implementation of the system is provided, demonstrating its functionality and
allowing comparison between different variants.

% The motivation for this thesis lies in understanding how mathematical problems
% from combinatorics and number theory can be used to construct cryptographic
% systems, as well as in exploring the challenges of balancing efficiency and
% security. By studying the Knapsack cryptosystem and its developments, this work
% contributes to a broader comprehension of public-key cryptography and its
% diverse mathematical foundations.

