\chapter{Theoretical Background}

\section{Basics of Cryptography} Cryptography is a part of computer science and
mathematics that focuses on protecting contents of transmitted data through
insecure channels like the Internet, from unauthorized parties. Its name is
derived from the greek word "kryptos" meaning hidden and "graphein" meaning
writing so we could literally translate it as hidden writing. In today's world
it is possible to hide any type of digital communication from text to video. 

The need for secret communication came much earlier. One of the first
implementations of cryptography was found in the use of non-standard
hieroglyphs carved into the wall of a tomb from the Old Kingdom of
Egypt~\cite{ibmCryptographyHistory}. This is an example of a classical cipher.
Classical ciphers came before the invention of computers and for that reason
they operate on letters rather than bits. In ancient times or even during the
First World War, cryptography had to be performed manually using only pen and
paper~\cite{aumasson2017serious}. The most famous classical cipher is Caesar
cipher. It is named after Julius Caesar, who used it during his rule of Roman
Empire to protect his military communications. It works by shifting each letter
in message by a number of positions in alphabet. The number of the shifted
letters must be agreed upon in advance between communicating parties and can be
considered as key. Even though this worked well in times of Roman Empire, it's
very easy to break. There are only 26 letters in English alphabet, hence only
26 keys possible. An attacker can simply try all possible shifts to quickly
recover the plaintext. This weakness later motivated the development of more
secure systems. 

Another popular classical cipher is the Vigenère cipher which came around 1500
years after the invention of Caesar cipher. The Vigenère cipher became popular
and was later used during the American Civil War by Confederate forces and
during the First World War by the Swiss Army, among
others~\cite{aumasson2017serious}. It works similarly to the Caesar cipher, but
rather than shifting every letter by a constant number of positions it shifts
by values defined by key. In this context, the key is a word in which each
letter represents numerical value derived from their position in alphabet.
During encryption, the key is repeated until it matches the length of the
message. Then, each letter of the message is shifted forward in the alphabet by
the value corresponding to the matching letter of the key. 

Despite the improvements, it's still quite easy to break with frequency
analysis techniques which exploit the uneven distribution of letters in
languages. Classical ciphers are insecure because they rely on basic operations
that can be done manually and modern computers can break them with basic
programs analysing letter frequencies or even brute force testing all possible
keys.

As the need for secure communications grew, more sophisticated systems were
developed, including mechanical rotor machines such as the Enigma and Lorenz
ciphers used during the Second World War. These devices implemented rapidly
changing substitution alphabets and, at the time, were considered as secure.

Their eventual breaking showed a primary flaw of classical cryptography.
Security of classical and mechanical ciphers depended on complicated
substitutions but they lacked computational hardness.

During the 1970s and 1980s cryptography progressed into a modern era. New
theory based on well-defined mathematical principles appeared, which led
researchers to realize that long-term security could not rely on secrecy or
complicated mechanical design, but instead on problems that are fundamentally
hard to solve. Modern cryptography is therefore built upon mathematically
defined functions that are easy to compute in one direction but infeasible to
reverse without secret information, called a key. These ideas led to the
development of both symmetric and asymmetric cryptosystems. The next section
introduces the computational principles underlying these constructions.

% To achieve this it relies on hard mathematical problems and algorithms such as
% integer factorization, discrete logarithms, or the subset-sum (knapsack)
% problem. These problems belong to a class known as NP-complete which means it
% is very difficult to find a solution within reasonable time without any
% information about input even with vast computational power. On the other hand
% verifying that the solution is correct is easy. All NP-complete problems are
% decision problems that for any input the output is yes or no. For that reason
% they form the foundation of modern cryptosystems. 
% 
% Because solving these problems directly is practically impossible, modern
% cryptography uses mathematical functions that depend on secret information
% called keys. With knowledge of the key the computation becomes simple, but
% without it the problem remains computationally infeasible. This idea sparked
% development of wide range of encryption methods and eventualy to both symmetric
% and asymetric cryptosystems.
\section{P, NP, and Computational Hardness}

In cryptography, the security of many systems depends on problems that are
believed to be computationally hard to solve. To describe meaning to
computational difficulty, problems are organized into complexity classes. The
groups are based on how the running time of an algorithm grows with the size of
the input. One important class is $\mathcal{P}$, which contains decision problems that
can be solved in polynomial time. A decision problem meaning one with a yes or
a no answer is in $\mathcal{P}$ if there exists an algorithm that solves any instance in
time bounded by a polynomial in the input size. Intuitively, problems in $\mathcal{P}$
are considered efficiently solvable.

Another fundamental class is $\mathcal{NP}$, which stands for nondeterministic
polynomial time. A decision problem belongs to $\mathcal{NP}$ if a proposed
solution, often called a certificate, can be verified in polynomial time.
This means that even if we do not know how to find a solution quickly, we can
check whether a given solution is correct in a reasonable time. 

% It is also
% important to note that $\mathcal{NP}$ does not mean “not polynomial”; instead,
% it refers to the way problems are defined in terms of verifiability.

A major open problem in theoretical computer science is whether
$\mathcal{P} = \mathcal{NP}$. Many researchers believe that $\mathcal{P} \neq
\mathcal{NP}$, but this has not been proven. If it turned out that
$\mathcal{P} = \mathcal{NP}$, then every problem whose solution can be verified
quickly could also be solved quickly. This would have enormous consequences:
many problems currently considered difficult would become easy, and most modern
cryptographic systems would no longer be secure. Because no polynomial-time
algorithm is known for many $\mathcal{NP}$ problems, the prevailing assumption
is that $\mathcal{P}$ $\neq$ $\mathcal{NP}$.

Within the class $\mathcal{NP}$  there exists a subset of problems that are, in a
precise sense, the most computationally difficult problems in the class. These
are referred to as $\mathcal{NP}$-complete problems. A decision problem $C$ is said to
be $\mathcal{NP}$-complete if it satisfies two requirements: it belongs to
$\mathcal{NP}$, and every problem in $\mathcal{NP}$ can be transformed into C by a
polynomial-time reduction. If a polynomial-time algorithm were discovered for
any $\mathcal{NP}$-complete problem, then all problems in $\mathcal{NP}$ would become
solvable in polynomial time. Since no such algorithm is known,
$\mathcal{NP}$-complete problems are generally regarded as computationally
unsolvable. 

One well-known $\mathcal{NP}$-complete problem is the subset-sum problem. In the decision
version of this problem, we are given a set of numbers and a target sum, and we
ask whether there exists a subset of the numbers whose sum equals the target.
It is easy to verify whether a chosen subset achieves the target sum, but
finding the correct subset appears to require exponential time in the worst
case. This contrast between easy verification and difficult search is typical
for $\mathcal{NP}$-complete problems and is one of the reasons why the Knapsack problem has
been used in cryptography.

There is also a broader category known as $\mathcal{NP}$-hard problems. Informally,
$\mathcal{NP}$-hard means that a problem is at least as difficult as any problem in
$\mathcal{NP}$. More formally, a problem $H$ is $\mathcal{NP}$-hard if every problem in
$\mathcal{NP}$ can be reduced to $H$ in polynomial time. $\mathcal{NP}$-hard problems do not
need to be decision problems, and they do not need to have efficiently
verifiable solutions. If an $\mathcal{NP}$-hard problem is also in $\mathcal{NP}$, then it
is $\mathcal{NP}$-complete. In other words, $\mathcal{NP}$-complete problems are exactly those that lie
in the intersection $\mathcal{NP} \cap \mathcal{NP}$-hard.

Modern cryptographic algorithms rely on the assumed hardness of such problems.
The security of a cryptosystem is based on the fact that no efficient algorithm
is known for solving its underlying mathematical problem. For instance, the RSA
cryptosystem is based on the difficulty of integer factorization, which is not
known to be $\mathcal{NP}$-complete but is still considered hard. The classical Merkle--Hellman
knapsack cryptosystem, on the other hand, is built directly on the $\mathcal{NP}$-complete
Knapsack problem. Because solving the general knapsack problem appears to
require exponential time, an attacker without the secret key faces an extremely
difficult computational task. As long as no efficient algorithms are found for
these kinds of hard problems, cryptographic schemes based on them remain secure.

\section{Symmetric and Asymmetric Cryptography}

Modern cryptography is divided into two main types. Symmetric and Asymmetric
also known as public-key cryptography. Aim of both is to secure communication,
but they differ in how the keys are used and mathematical concepts that make
them work giving each approach its own strengths and weaknesses.

Symmetric cryptography follows the same basic idea that classical ciphers were
built on. In this approach, all parties involved in transaction use a single
shared secret key for both encryption and decryption. They must exchange this
key in a secure way before communicating, so key distribution becomes an
important problem.

Because of this limitation, symmetric systems are often not practical when two
parties have never communicated before or do not have a secure channel for
exchanging the key. This problem motivated the development of a different
approach, where the communicating parties do not need to share a secret in
advance. This idea led to asymmetric, or public-key cryptography.

Asymmetric cryptography works with a pair of keys: a public key that can be
shared with anyone and a private key that must remain secret. Messages
encrypted with the public key can only be decrypted with the corresponding
private key, which removes the need for a secure key exchange. The security of
these systems depends on mathematical problems that are easy to compute in one
direction but extremely difficult to reverse without the private key.
...
These ideas form the basis of modern public-key cryptography. Many encryption
schemes rely on specific mathematical problems that are assumed to be hard to
solve without the secret key. One such problem is the subset-sum, also known as
the knapsack problem, which plays a central role in the cryptosystem studied in
this thesis. The following sections introduce this problem and explain how it
can be used to construct a public-key encryption scheme.

\section{The Knapsack Problem}

The knapsack problem, also known as the subset--sum problem, is a fundamental
combinatorial decision problem. Given a sequence of positive integers
\[
W = (w_1, w_2, \dots, w_n)
\]
and a target value \(s \in \mathbb{N}\), the task is to determine whether there exists
a binary vector
\[
x = (x_1, x_2, \dots, x_n) \in \{0,1\}^n
\]
such that
\[
\sum_{i=1}^{n} x_i w_i = s.
\]
In this formulation, each variable \(x_i\) indicates whether the weight \(w_i\) is
included in the subset. The problem is NP--complete: verifying a proposed solution
is efficient, but no polynomial--time algorithm is known for solving arbitrary
instances.

\subsection*{Example of a general instance}

Consider the sequence
\[
W = (3,\, 8,\, 12,\, 19)
\]
and the target value
\[
s = 27.
\]
To decide whether a subset sums to \(27\), one must explore combinations of the
weights. In this case, a valid solution exists:
\[
8 + 19 = 27,
\]
corresponding to the vector
\[
x = (0,\,1,\,0,\,1).
\]
For a general instance, systematically finding such a vector may require testing
an exponential number of subsets in the worst case.

\subsection*{Superincreasing sequences}

A sequence \(W\) is called \emph{superincreasing} if every element is strictly larger
than the sum of all previous ones:
\[
w_i > \sum_{j=1}^{i-1} w_j \qquad \forall i \ge 2.
\]
For such sequences, the subset--sum problem is solvable by a simple greedy algorithm
that starts with the largest weight and works downward.

For a general subset-sum instance with an arbitrary sequence of weights, the
target value may have no solution, one solution, or multiple different
solutions. More importantly, without additional structure, finding a valid
subset may require exponential search in the worst case. In contrast,
superincreasing sequences have strong structure: each decision in the greedy
process is forced, so the solution (if it exists) is recovered efficiently and
deterministically. This property is the key reason why superincreasing
sequences are used as a private trapdoor in the classical Merkle--Hellman
construction.

\subsection*{Example of a superincreasing instance}

Consider the sequence
\[
W = (2,\, 5,\, 11,\, 23),
\]
which satisfies
\[
5 > 2,\quad 11 > 2+5,\quad 23 > 2+5+11.
\]
Let the target value be
\[
s = 28.
\]
The greedy procedure proceeds as follows:
\begin{enumerate}
    \item \(23 \le s\), include it: \(x_4 = 1\), new \(s = 5\).
    \item \(11 > 5\), skip it: \(x_3 = 0\).
    \item \(5 \le 5\), include it: \(x_2 = 1\), new \(s = 0\).
    \item \(2 > 0\), skip it: \(x_1 = 0\).
\end{enumerate}
The algorithm terminates with the unique solution
\[
x = (0,\,1,\,0,\,1),
\]
corresponding to
\[
23 + 5 = 28.
\]

\subsection*{Summary}

The essential feature of the knapsack problem is the strong contrast between
general instances, which appear computationally intractable, and structured
instances such as superincreasing sequences, which admit efficient solutions.
This distinction forms the mathematical basis for constructions discussed in
later chapters.

\section{Knapsack Cryptosystem}

The Merkle--Hellman cryptosystem was one of the earliest public-key encryption
schemes based on the subset-sum problem \cite{MerkleHellman1978}. It was
introduced in a model where secure communication does not require pre-shared
secret keys. Its design uses a superincreasing private sequence as a trapdoor
and publishes a transformed sequence that should appear computationally hard to
solve without the private key. Although the idea was influential, later
cryptanalysis showed that the basic construction is insecure \cite{Shamir1982}.
The following description focuses on the classical Merkle--Hellman construction
itself, namely key generation, encryption, and decryption.

\subsection{Key Generation}

The goal of key generation is to create two representations of the same
knapsack instance: one easy for the owner and one hard-looking for everyone
else \cite{MerkleHellman1978}. The process starts by choosing a superincreasing
sequence \(W=(w_1,\dots,w_n)\), which will serve as the private trapdoor. Next,
the key owner chooses a modulus \(m\) larger than the sum of all private
weights and a multiplier \(r\) that is invertible modulo \(m\). These values
are used
to hide the private sequence by modular transformation, producing a public
sequence \(B=(b_1,\dots,b_n)\), where \[ b_i \equiv r\,w_i \pmod m. \] 

In the classical description, a permutation is also applied to further obscure
the original structure. The public key is the transformed sequence (and
permutation, if used), while the private key consists of the superincreasing
sequence and transformation parameters \cite{MerkleHellman1978}.

\subsection{Encryption}
After key generation, the sender uses the public sequence \[
B=(b_1,b_2,\dots,b_n) \] to encrypt an \(n\)-bit message \[
x=(x_1,x_2,\dots,x_n), \quad x_i \in \{0,1\}. \] The ciphertext is computed
as \[ C=\sum_{i=1}^{n} x_i b_i. \] This means that only the public weights
corresponding to message bits equal to 1 are added. From the sender's
perspective, encryption is therefore straightforward and efficient, because
it only requires additions over public values \cite{MerkleHellman1978}.

\subsection{Decryption}
To decrypt the ciphertext \(C\), the key owner uses private parameters and
first computes
\[
S' \equiv r^{-1} C \pmod m,
\]
where \(r^{-1}\) is the modular inverse of \(r\) modulo \(m\). Because the
public weights were constructed from the private superincreasing sequence,
this step maps the problem back to an equivalent subset-sum instance over
the private sequence. The resulting value \(S'\) is then solved by a greedy
procedure, starting from the largest private weight and moving downward. At
each step, if the current weight does not exceed the remaining sum, the
corresponding bit is set to 1 and the weight is subtracted; otherwise the
bit is set to 0. If permutation is used during key generation, bit positions
are restored according to the secret key order after greedy recovery
\cite{MerkleHellman1978}.

\subsection*{Example}
Consider a private superincreasing sequence
\[
W=(2,\,5,\,11,\,23),
\]
with \(m=47\) and \(r=9\), where \(\gcd(r,m)=1\) and \(m>\sum W=41\).
The public key is
\[
B=(18,\,45,\,5,\,19),
\]
computed as \(b_i \equiv r w_i \pmod m\).
For message \(x=(1,0,1,0)\), encryption gives
\[
C=18+5=23.
\]
For decryption, \(r^{-1}\equiv 21 \pmod{47}\), so
\[
S' \equiv r^{-1}C \equiv 21\cdot 23 \equiv 13 \pmod{47}.
\]
Using greedy recovery on \(W\): \(13=11+2\), so recovered bits are \(x=(1,0,1,0)\).

\subsection*{Assumptions and Limitations}
Although the classical Merkle--Hellman design is efficient for encryption and
decryption, it assumes that the transformation from the private
superincreasing sequence to the public sequence sufficiently hides the trapdoor
structure. In practice, this assumption is not strong enough for the basic
construction: later cryptanalysis showed that equivalent private structure can
be recovered efficiently in polynomial time \cite{Shamir1982}. Therefore, the
original scheme is historically important, but not secure for real-world use,
which motivates the analysis of weaknesses and attacks in the next chapter.
